---
title: "Make Signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Make Signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The bugphyzz package provides the `makeSignatures` function, which allows to
create lists of microbe signatures per attribute from a dataset imported with
the `physiologies` or `fattyAcidComposition` functions. These lists of
signatures can be used for enrichment analysis using other R packages,
such as `EnrichmentBrowser`.

```{r setup}
library(bugphyzz)
library(dplyr)
```

# Import bugphyzz datasets

The first step to crate a list of signatures is to import a bugphyzz dataset.
This dataset can have attributes with categorical (e.g. aerophilicity) or 
continuous values (e.g. growth temperature).

A dataset with attributes with categorical values:

```{r}
aer <- physiologies("aerophilicity")[[1]]
glimpse(aer)
```

An dataset with an attribute with continuous values:

```{r}
gt <- physiologies("growth temperature")[[1]]
glimpse(gt)
```

# Make signatures

The `makeSignatures` function allows to create signatures at different
taxonomic ranks and with different taxids.

Signatures with NCBI taxids at the genus level:

```{r}
aer_sig <- makeSignatures(aer, taxids = "NCBI_ID", tax_rank = "genus")
lapply(aer_sig, head)
```

Signatures with taxon names at the strain level:

```{r}
aer_sig <- makeSignatures(aer, taxids = "Taxon_name", tax_rank = "strain")
lapply(aer_sig, head)
```

## Inherirted signatures

The taxa annotated in bugphyzz belongs to different taxonomic ranks and from
different sources. This could lead to some taxa annotated in a dataset at a
rather low taxonomic level, e.g strain, but not at a higher taxonomic level, 
e.g. species. The implementation of inherited signatures allows that a given
taxa inherits an attribute annotation from a lower rank taxa if they belong
to the same lineage. This could lead to an increase of represented taxa per
attribute.

For example, let's create a signature of thermophilic bacteria, i.e. with
a growth temperature above 80 °C.

The first step is to filter the taxa with a growth temperature above 80 °C:

```{r, fig.width=5, fig.height=5}
gt_thermophilic <- gt %>% 
  filter(Attribute_value > 80)

hist(gt_thermophilic$Attribute_value, 
     xlab = "Growth temperature (°C)", main = NULL)
```


Now we can use the `makeSigature` function to create a microbe signature at
the species level using the taxa names:

```{r}
gt_therm_sig <- makeSignatures(gt_thermophilic, taxids = "Taxon_name",
                               tax_rank = "species")
gt_therm_sig
```

In the output above there are only 15 species. However, if we set the
`inherit` argument to `TRUE`, we can increase the number of thermophilic 
species represented in our microbe signature:

```{r}
gt_therm_sig_inherited <- makeSignatures(gt_thermophilic, taxids = "Taxon_name",
                                         tax_rank = "species", inherited = TRUE)
gt_therm_sig_inherited
```

The number of species represented in our signature reached 35.

We recommend caution in the use and interpretation of the inherited signatures
option. For example, the output in the code below indicates that there
are one or more thermophilic taxa representing the Archaea and Bacteria 
superkingdoms in the 'growth temperature' dataset, 
not that all Arcahea and Bacteria taxa are thermophilic.

```{r}
makeSignatures(gt_thermophilic, taxids = "Taxon_name", 
               tax_rank = "superkingdom", inherited = TRUE,
               min_sig_size = 1)
```

