---
title: "bugphyzz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bugphyzz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

This document shows the main functions and operations that can be used
to work with the bugphyzz datasets, such as seeing all physiologies, 
look at specific physiologies, and make signature sets. _bugphyzz_ is a collection of physiological features about bacteria. We are curating the databases through 3 main methods: webscraping, bulk download from online databases, and manual curation from literature. Mainly this database will be used to analyze microbiome data.

## Definitions

### Always present

* **NCBI ID**: National Center for Bioinfomatic numeric ID to link to specific taxa.
* **Taxon name**: a human readable organismal name that can delinate which level taxa the annotation is describing.
* **Attribute**: a value for a physiology, that is mapped to an ontology whenever possible. These are defined in the `attribute` column of [inst/extdata/attributes.tsv](https://github.com/waldronlab/bugphyzz/blob/main/inst/extdata/attributes.tsv
 **Physiology**: a microbial characteristic that can be measured or observed, that should be mapped to an ontology wherever possible. These are defined in [inst/extdata/links.tsv](https://github.com/waldronlab/bugphyzz/blob/main/inst/extdata/links.tsv).
* **Attribute value**: a value describing a physiology. Allowable values are defined by the `validity` column of [inst/extdata/attributes.tsv](https://github.com/waldronlab/bugphyzz/blob/main/inst/extdata/attributes.tsv). For categorical attributes (such as "obligately aerobic"), values can be "TRUE" and "FALSE", for example. For numeric attributes, these can be numeric values.
* **Attribute source**: a citeable domain which the annotation was obtained. A source can be another database or primary literature.
* **Frequency**: Describes how often the attritbute (phenotype) occurs: Rarely, Sometimes, Usually, Always, or Unknown. If the source makes the attribute unclear how often it occurs, then an unknown is given.
* **Evidence codes**: As the Gene Ontology defines different forms of evidence a source was obtained the annotation. 
    - EXP: Experimental data, the attribute was obtained through labortory confirmations, and ideally more than once.
    - IGC: Computational predicted through means of metagenomic analysis, or more broadly, genome analysis.
    - ASR: Anstrestal State Reconstruction, was predicted using ASR computional techniques to predicted up and down the taxonmic tree of taxa that has a value.
* **Note**: Any additional information that does not fit in any fixed column. The information here is variable.
* **Parent name**: The parent of descendant given in Taxon name column. Values can includes names or NA.
* **Species name**: The species name, or descendant of parent, given in the Taxon name column. Values can includes names or NA.
* **Unit**: defines unit of attribute value, like µm.
* **Rank**: The taxonomic rank of the annotation. Values can includes names or NA.
* **Parent Rank**: The taxonmic rank of the parent of the annotation. Values can includes names or NA.

### Sometimes present

* **MIDAS ID**: Located in the Habitat attribute. Any annotation from the MIDAS database is assigned an ID number. We provided that ID here.
* **Genome ID**: From NCBI, another type of numeric ID, that links to a specific sequenced genome. Not all taxa have a NCBI ID, and might only have a Genome ID
* **Accession ID**: The accession number is a unique identifier assigned to a record in sequence databases such as GenBank. Several NCBI databases use the format `[alphabetical prefix][series of digits]`. A change in the record in some databases (e.g. GenBank) is tracked by an integer extension of the accession number, an Accession.

## bugphyzzAnalyses

Some analyses using data provided by this package can be found at https://waldronlab.io/bugphyzzAnalyses/

## How to add a new attribute in GitHub to the BugPhyzz database

Curation steps:
 1. Add attributes of physiology to Attributes using ontology terms
 2. Add reference(s) to Confidence in Curation

If using Google Sheets, publish a csv file to web:
 4. Add Google Doc "publish to web" link and add in "Link" https://github.com/waldronlab/bugphyzz/blob/main/inst/extdata/links.tsv 
 5. Add Google Doc link to "Source Link" https://github.com/waldronlab/bugphyzz/blob/main/inst/extdata/source_links.tsv

Steps in the `bugphyzz` package:
 1. _Samuel fill in_

## Loading datasets in R

```{r, message=FALSE, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("waldronlab/bugphyzz")
```

```{r setup, message=FALSE}
library(bugphyzz)
library(purrr)
library(tibble)
```

The main function for importing datasets is the `physiologies` function, which
has a single argument, `keyword`, and returns a list of data.frames.
The `keyword` argument accepts a character vector specifying the name(s) 
of the dataset(s) that will be imported.

A datset with categorical attribute values:

```{r, message=FALSE}
aer <- as_tibble(physiologies("aerophilicity")[[1]])
aer
```

A dataset with numeric attribute values:

```{r}
ph <- as_tibble(physiologies('optimal ph')[[1]])
ph
```

The `physiologiesList` function can be used to display which datasets are
available through the `physiologies` function before importing them:

```{r}
showPhys()
```

You can use the `keyword = "all"` to import all of the datasets:

```{r, message=FALSE, eval=FALSE}
## Import all of the datasets available through the physiologies function
phys <- map(physiologies(keyword = "all"), as_tibble)
```

A separate function is provided to import the fatty acid compositions dataset,
`fattyAcidComposition`. This function takes no arguments and returns a single
data.frame object:

```{r}
fac <- as_tibble(fattyAcidComposition())
fac
```

## Some examples of how to create signatures

A signature of attributes with character values:

```{r, message=FALSE}
aer_sig <- makeSignatures(
    keyword = 'aerophilicity', tax.id.type = 'NCBI_ID', tax.level = 'genus', 
    Frequency = 'always', Evidence = 'exp', min.size = 10
)

map(aer_sig, head)
```

A signature of an attribute with numeric values (acidic bacteria):

```{r}
acid_sig <- makeSignatures(
    keyword = 'optimal ph', tax.id.type = 'Taxon_name', tax.level = 'species',
    max = 4
)
map(acid_sig, head)
```

## Session information:

```{r}
sessioninfo::session_info()
```
