---
title: "bugphyzz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bugphyzz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bugphyzz)
```

## Introduction

This document shows the main functions and operations that can be used
to work with the bugphyzz datasets.

## Installation

Bugphyzz can be installed with:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager") 

BiocManager::install("waldronlab/bugphyzz")
```


## Importing datasets

The main function for importing datasets is the `physiologies` function, which
has a single argument, `keyword`, and returns a list of data.frames.
The `keyword` argument accepts a character vector specifying the name(s) 
of the dataset(s) that will be imported:

```{r import aerophilicity, message=FALSE}
## Import a single dataset
aer <- physiologies("aerophilicity")[[1]]
head(aer)
```

You can use the `keyword = "all"` to import all of the datasets:

```{r import physiologies, message=FALSE}
## Import all of the datasets available through the physiologies function
phys <- physiologies(keyword = "all")
str(phys, max.level = 1)
```

The `physiologiesList` function can be used to display which datasets are
available through the `physiologies` function before importing them:

```{r physiologiesList}
physiologiesList()
```

A separate function is provided to import the fatty acid compositions dataset,
`fattyAcidComposition`. This function takes no arguments and returns a single
data.frame object:

```{r import fac}
fac <- fattyAcidComposition()
head(fac)
```

## Creating microbial signatures

Bugphyzz provides the `makeSignatures` function to create a list of microbial
signatures from a bugphyzz dataset.

```{r aerophilicity signature}
aer_sig <- makeSignatures(
  aer, tax_id_type = "Taxon_name", tax_level = "genus", evidence = "EXP",
  ci = 1, min_size = 10
)
lapply(aer_sig, head)
```

In some cases, some filtering might be necessary before creating the signatures.
For example, let's create a signature of microbes that grow above 80째C.

First, we must import the "growth temperature" dataset.

```{r import growth temperature, message=FALSE}
gt <- physiologies("growth temperature")[[1]]
head(gt)
```

Now, we filter only those taxa that grow above 80째C with regular R functions:

```{r filter termophiles, fig.height=5, fig.width=5}
gt_subset <- gt[gt$Attribute_value >= 80, ]
hist(
  x = gt_subset$Attribute_value, xlab = "Temperature (째C)", 
  main = "Microbes that grow above 80 째C"
)
```

Finally, we can create the microbial signatures from the filtered dataset.
In his case, a signature of species using only the taxids.

```{r thermophiles signature}
gt_subset_sig <- makeSignatures(
  gt_subset, tax_id_type = "NCBI_ID", tax_level = "species",
  evidence = c("Unknown", "EXP", "ASR"), ci = 0.4, include_unknown_ci = TRUE
)
head(gt_subset_sig)
```

Session information:

```{r sessionInfo}
sessionInfo()
```

