---
title: "Example of a Signature Score for a Categorical Attribute (aerophilicity)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example of Signature Score for a Categorical Attribute (aerophilicity)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r , include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning=FALSE
)
```

```{r setup, message=FALSE, warning=FALSE}
library(bugphyzz)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(ComplexHeatmap)
```


Example using the **aerophilicity** data set

```{r}
aerophilicity <- physiologies("aerophilicity")[[1]]
aerophilicity[ , c("NCBI_ID", "Attribute")] %>% 
  head() %>% 
  knitr::kable()
```

For this example, the AsnicarF_2017 metaphlan bugs lists are used, which have been filtered so that they only contain the relative abundance of the bacteria at the genus level.

The metaphlan bugs list of each of the 26 samples from the AsnicarF_2017 data set are provided as separate data frames in `AsnicarF_2017_genus`. Each data frame contains two columns: "NCBI_ID" (NCBI TaxIDs) and "Abundance."

```{r}
str(AsnicarF2017_genus, max.level = 1)
AsnicarF2017_genus[[1]] %>% 
  knitr::kable()
```

NCBI taxids can be matched between a data frame in `AsnicarF2017_genus` and the NCBI_ID column in the cellular respiration data set ("aerophilicity") of the bugphyzz package. For example, for the first sample in `AsnicarF2017_genus`:

```{r}

sample1_taxids <- AsnicarF2017_genus[[1]]$NCBI_ID
aerophilicity_taxids <- aerophilicity$NCBI_ID

AsnicarF2017_genus[[1]][sample1_taxids %in% aerophilicity_taxids, ] %>% 
  knitr::kable()
```

In the example above, only 12 taxids in the `AsnicarF2017_genus[[1]]` data frame matched a taxid in the aerophilicity database.

We can use the `add_attributes` function to add the attributes as columns to each data frame. The inputs are a data frame (containing taxids and relative abundance data) and the aerophilicity data set from bugphyzz (containing taxids and attributes).

Moreover, the `add_hierarchy` function, which implements the `taxize::classification` function, can be used to add a column with the genus name of each taxid.

The output is the same metaphlan bugs list with added attribute columns--one per attribute The values in the new columns match the abundance values for each genus.

```{r, message=FALSE, warning=FALSE}
x <- map(AsnicarF2017_genus, ~add_attributes(.x, aerophilicity))
df <- x[[1]] %>% 
  add_hierarchy("genus") 
df %>% 
  knitr::kable()

```

To get the signature scores for a single sample, relative abundance values per attribute column are added up, reflecting the total relative abundance of bacteria per attribute.

Notice that the score cannot be higher than 100 for each attribute, but if a single genus have more than one attribute in the bugphyzz database, its abundance value is added up in two different columns.
 
In the example below, in a single sample 84% of bacteria are areobic and 67% are facultative anaerobic.

```{r}
x_score <- get_score(x[[1]])
x_score %>% 
  knitr::kable()
```

This can be applied to all the samples (data frames) of the `AsnicarF2017_genus` data set, getting a singature per sample:

```{r}
y <- map(x, ~get_score(.x))
str(y, max.level = 1)
```

And the scores of all samples can be merged into a single matrix:

```{r, message=FALSE, warning=FALSE}
for (i in seq_along(y)) {
    names(y[[i]])[[2]] <- names(y)[i]
}

df <- purrr::reduce(y, full_join)
mat <- as.matrix(df[,-1])
mat[is.na(mat)] <- 0
rownames(mat) <- df$Attribute

mat[, 1:5] %>% 
  knitr::kable()
```

Create a heatmap:

```{r, message=FALSE, warning=FALSE, fig.align='center', fig.dim=c(10,5)}

mat <- mat[rowSums(mat) >= 1, ]

ht1 <- ComplexHeatmap::Heatmap(mat, col = viridis::viridis(1024))
# See heatmap below
```

Merge the relative abundance of each genus of bacteria of all samples  into a single data frame:


```{r, warning=FALSE, message=FALSE}
output <- vector("list", length(AsnicarF2017_genus))

for (i in seq_along(output)) {
    
    output[[i]] <- AsnicarF2017_genus[[i]]
    names(output[[i]])[[2]] <- names(AsnicarF2017_genus[i])

}

abundance <- purrr::reduce(output, dplyr::full_join) %>%
    add_hierarchy("genus")

abundance <- abundance[abundance$NCBI_ID %in% aerophilicity_taxids, ]

abundance[1:10,1:5] %>% 
  knitr::kable()
```

Create a matrix of relative abundance per genus of all samples:

```{r}
abundance_matrix <- as.matrix(abundance[,-c(1,2)])
abundance_matrix[is.na(abundance_matrix)] <- 0
rownames(abundance_matrix) <- abundance$genus

head(abundance_matrix[,1:5]) %>% 
  knitr::kable()
```

Create a heatmap

```{r,warning=FALSE, message=FALSE, fig.dim=c(10,15)}

abundance_matrix <- abundance_matrix[ rowSums(abundance_matrix) > 8,]
ht2 <- Heatmap(abundance_matrix, col = viridis::viridis(1024),
              cluster_columns = FALSE)
```

Visualize both heatmaps:

```{r, message=FALSE, warning=FALSE, fig.dim=c(10,10)}
ht_list = ht1 %v% ht2
draw(ht_list)
```

# Application to HMP 2012 16S dataset

```{r}
suppressPackageStartupMessages(library(HMP16SData))
v13dat <- V13()
dim(v13dat)
```

Sum OTUs by genus
```{r}

v13assay <- assay(v13dat) %>%
  as_tibble()
v13assay$GENUS <- rowData(v13dat)$GENUS
v13asay <- v13assay %>% 
  relocate(GENUS, everything())

v13summed <- v13assay %>% 
  filter(!is.na(GENUS)) %>%
  group_by(GENUS) %>%
  summarise_all(sum)
dim(v13summed)
```

