---
title: "Curation Checks"
output:
  html_document:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(bugphyzz)
library(dplyr)
library(purrr)
```

# Import bugphyzz data (physiologies)

```{r, message=FALSE}
phys <- physiologies()
```

# Check presence and order of 'required' columns

The function `.checkRequiredColumnList()` performs two checks:  
1. All of the required columns are present.  
2. The required columns are in the right order.

```{r, message=FALSE}
err1 <- bugphyzz:::.checkRequiredColumnsList(phys, table = TRUE)
```

The result has four columns:  
1. **dataset** contains the name of each dataset.  
2. **error_type** contains the type of error ('missing column' or 'misplaced
column').  
3. **columns** the name of the columns that must be corrected.  
4. **link** the link to the dataset.

Visualization of the errors:

```{r}
if (!is.null(err1)) {
  DT::datatable(err1)
} else {
  message('No errors found.')
}
```

# Check syntax

`.checkColumnValuesList` performs two checks on the list of physiologies:
1. That a column is in the template file.  
2. That all of the values in a column are valid according to the template  
file.

```{r, message=FALSE}
phys2 <-  phys |> 
  map(~ dplyr::filter(.x, Attribute_source != 'https://bacdive.dsmz.de/')) |> 
  discard(~!nrow(.x))
err2 <- bugphyzz:::.checkColumnValuesList(phys2, table = TRUE)
```

## Uncatalogued columns

Columns in the output:

+ **dataset** the name of the spreadsheet.
+ **col** the name of the column that has not been added to the template.csv
file.
+ **source_link** link to the spreadsheet.

```{r}
uncat_columns <- dplyr::filter(err2, error_type == 'Uncatalogued column') |> 
  discard(~ all(is.na(.x))) |> 
  select(-error_type)
DT::datatable(uncat_columns)
```


## Errors in syntax/values

Columns in the output:

+ **dataset.** Name of the spreadsheet.
+ **col.**  Name of the column with invalid values.
+ **n_rows.** Number of rows with the same invalid value.
+ **invalid_values.** The value that most be corrected or added to
the attributes.tsv file or the template.tsv file
+ **source_link.** The link to the spreadsheet.

```{r, message=FALSE}
invalid_values <- dplyr::filter(err2, error_type == 'Invalid values') |> 
  discard(~ all(is.na(.x))) |> 
  tidyr::unnest(cols = c(invalid_values, invalid_pos)) |> 
  select(-error_type, -invalid_pos)
invalid_values |> 
  # head(100) |> 
  unique() |> 
  select(-col, -n_rows) |> 
  filter(dataset != 'isolation site') |> 
  DT::datatable(
    filter = 'top',
    extensions = c("Buttons","KeyTable"),
    options = list(
      dom = '<"top"B>lfrtip',
      buttons = list('copy', 'print'),
      iDisplayLength = 25,
      keys = TRUE,
      pageLength = 100, 
      scrollY = TRUE
    )
  )
```


# Check valid NCBI IDs

NCBI IDs are compared against the latest version of the NCBI ID database
available through `taxizedb`.

If an NCBI ID is **outdated** (some IDs are no longer in use or are merged into
another one), then a more recent NCBI ID and taxon name are suggested based on
the result of `taxize::classification`. You should see "outdated" in the 
comment column (below).

If no NCBI ID is suggested, it's likely because the taxid doesn't exist.
You should see "not found" in the comment column (below).

Note: taxize and taxizedb are not based on the same annotations. Both former
and updated annotations can be accessed through taxize. However, this can take
a long time since the client must make requests to the NCBI API. Having
an NCBI API key can help (it's free once you created your NCBI account). On the
other hand, working with the taxizedb functions is much faster. This is because
the NCBI database is downloaded first to the user's machine. Only the most
recent version of the database is downloaded, that's why only current 
annotations are show. 
Roughly, taxize behaves like https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi
while taxizedb behaves like https://www.ncbi.nlm.nih.gov/taxonomy/.


```{r, message=FALSE, warning=FALSE, eval=FALSE}
id_err <- lapply(phys, bugphyzz:::.check_valid_ncbi_ids) %>% 
  purrr::discard(is.null) %>%
  bind_rows(.id = 'dataset')

links_file <- system.file('extdata/links.tsv', package = 'bugphyzz')
links <- readr::read_tsv(links_file, show_col_types = FALSE)
links <- links[,c('physiology', 'source_link')]

id_err_tbl <- dplyr::left_join(
  id_err, links, by = c('dataset' = 'physiology')
)
dim(id_err_tbl)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
DT::datatable(id_err_tbl)
```


