---
title: "Descriptive statistcs of taxa annotations"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(bugphyzz)
library(dplyr)
library(purrr)
library(ggplot2)

```

```{r, message=FALSE}
phys <- physiologies(keyword = "all")
phys[["fatty acid composition"]] <- fattyAcidComposition()
taxonomy_table <- bugphyzz:::taxonomyAnnotations
## remove prefix
colnames(taxonomy_table) <- sub("bugphyzz_", "", colnames(taxonomy_table)) 
## Select only columns with rank names, not rank ids
taxonomy_table <- select(taxonomy_table, -ends_with("_id", ignore.case = FALSE))
taxonomy_table$NCBI_ID <- as.integer(taxonomy_table$NCBI_ID)
```

```{r}
taxonomy_table %>% 
    count(rank) %>% 
    arrange(-n)
```

# Taxa annotated with NCBI IDs by rank

```{r}
data <- phys %>% 
  map(~ select(.x, all_of(c("NCBI_ID", "Rank")))) %>% 
  map(~ modify_at(.x, "NCBI_ID", as.integer)) %>% 
  bind_rows(.id = "dataset") %>% 
  as_tibble()
```

```{r, fig.width=9, fig.height=10}
data %>% 
  mutate(Rank = forcats::fct_relevel(Rank, "genus", "strain", "species")) %>% 
  ggplot(aes(dataset)) +
  geom_bar(aes(fill = Rank), position = "fill") +
  scale_y_continuous(label = scales::percent) +
  labs(x = "Dataset", y = "NCBI_ID",
       title = "Taxa Annotations with NCBI IDs in Bugphyzz",
       subtitle = "physiologies + fattyAcidComposition") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "bottom", plot.margin = unit(c(1,2,1,1),"cm"))
```

# Functions 

```{r}
## Create summary statistics per dataset
attributeSummaryByRank <- function(x, y, categorical = TRUE) {
    
    if (isTRUE(categorical)) {
        df <- x[[y]] %>%
            dplyr::filter(Attribute_value == TRUE) 
    } else if (isFALSE(categorical)) {
        df <- x[[y]]
    }
    
    df %>%
        dplyr::select(Attribute, Rank, NCBI_ID) %>%
        dplyr::mutate(
            rank = forcats::fct_relevel(Rank, "genus", "strain", "species"),
            Attribute = stringr::str_squish(Attribute)
        ) %>% 
        dplyr::count(Attribute, rank) %>% 
        dplyr::group_by(Attribute) %>% 
        dplyr::mutate(
            percent = round(n / sum(n) * 100, 2),
           `n (%)` = paste0(n, " (", percent, ")")
        ) %>% 
        dplyr::ungroup() %>%
        dplyr::select(Attribute, rank, `n (%)`) %>% 
        dplyr::arrange(Attribute, rank) %>% 
        tidyr::pivot_wider(names_from = "rank", values_from = `n (%)`) %>% 
        as.data.frame() %>% 
        knitr::kable(row.names = TRUE)
}

## Create histogram (for datasets with continuous values)
display_freq <- function(data, phys) {
    data %>%
        ggplot2::ggplot(ggplot2::aes(Attribute_value)) +
        ggplot2::geom_histogram() +
        ggplot2::labs(x = phys, y = "Frequency", title = phys) +
        ggplot2::theme_bw()
}

## Create boxplot per phylum (for datasets with continuous values)
boxplot_per_phylum <- function(data, phys, unit = "unit") {
    title <- paste(phys, "per phylum")
    y <- paste(phys, "(", unit, ")")
    data %>%
       ggplot2::ggplot(ggplot2::aes(phylum, Attribute_value)) +
       ggplot2::geom_boxplot() +
       ggplot2::labs(title = title, y = y, x = "Phylum") +
       ggplot2::theme_bw() +
       ggplot2::theme(
         axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
       )
}

```

# Datasets with categorical attributes in "physiologies"

```{r categorical attributes}
taxonomy_table$NCBI_ID <- as.character(taxonomy_table$NCBI_ID)
categorical_phys <- phys %>%
    keep(~is.logical(.x$Attribute_value)) %>% 
    map( ~ left_join( .x, taxonomy_table, by = "NCBI_ID"))
categorical_phys_names <- names(categorical_phys)
categorical_phys_names
```

```{r, warning=FALSE}
cat_summary_tables <- categorical_phys_names %>% 
    map(~ attributeSummaryByRank(categorical_phys, .x)) %>% 
    set_names(categorical_phys_names)
```

## Animal pathogen

```{r}
cat_summary_tables$`animal pathogen`
```

## Antimicrobial resistance

```{r}
cat_summary_tables$`antimicrobial resistance`
```

## Antimicrobial sensitivity

```{r}
cat_summary_tables$`antimicrobial sensitivity`
```

## Biofilm forming

```{r}
cat_summary_tables$`biofilm forming`
```

## Arrangement

```{r}
cat_summary_tables$arrangement
```

## Shape

```{r}
cat_summary_tables$shape
```

## Extreme environment

```{r}
cat_summary_tables$`extreme environment`
```

## Gram stain 

```{r}
cat_summary_tables$`gram stain`
```

## Growth medium

```{r}
cat_summary_tables$`growth medium`
```

## Habitat

```{r}
cat_summary_tables$habitat
```

## Aerophilicity

```{r}
cat_summary_tables$aerophilicity
```

## Plant pathogenicity

```{r}
cat_summary_tables$`plant pathogenicity`
```

## Spore shape

```{r}
cat_summary_tables$`spore shape`
```

## Isolation site

```{r}
cat_summary_tables$`isolation site`
```
## Disease association

```{r}
cat_summary_tables$`disease association`
```


# Datasests with numeric attributes in "physiologies"

```{r}
phys$`fatty acid composition` <- phys$`fatty acid composition` |> 
  mutate(NCBI_ID = as.character(NCBI_ID))
numeric_phys <- phys %>% 
    keep(~is.numeric(.x$Attribute_value)) %>% 
    map(~left_join(.x, taxonomy_table, by = "NCBI_ID"))

numeric_phys_names <- names(numeric_phys)

numeric_phys_names
```

```{r}
numeric_summary_tables = map(numeric_phys_names, ~ attributeSummaryByRank(numeric_phys, .x, categorical = FALSE)) %>% 
    set_names(numeric_phys_names)

histograms <- map2(
  phys[numeric_phys_names], numeric_phys_names, ~ display_freq(.x, .y)
)

boxplots <- map2(
  phys[numeric_phys_names], numeric_phys_names, ~ {
    x <- left_join(.x, taxonomy_table, by = "NCBI_ID")
    boxplot_per_phylum(x, .y)
  }
)
```

## Butyrate producing

```{r}
numeric_summary_tables$`butyrate producing`
```

```{r}
histograms$`butyrate producing`
```

```{r}
boxplots$`butyrate producing`
```

## Mutation rate per site per generation

```{r}
numeric_summary_tables$`mutation rate per site per generation`
```

```{r}
histograms$`mutation rate per site per generation`
```

```{r}
boxplots$`mutation rate per site per generation`
```

## Mutation rates per site per year

```{r}
numeric_summary_tables$`mutation rates per site per year`
```

```{r}
histograms$`mutation rates per site per year`
```

```{r}
boxplots$`mutation rates per site per year`
```

## Growth temperature

```{r}
numeric_summary_tables$`growth temperature`
```

```{r, message=FALSE}
histograms$`growth temperature`
```

```{r}
boxplots$`growth temperature`
```

## Optimal pH in "physiologies"

```{r}
numeric_summary_tables$`optimal ph`
```

```{r}
histograms$`optimal ph`
```

```{r}
boxplots$`optimal ph`
```

## COGEM pathogenicity rating

```{r}
numeric_summary_tables$`COGEM pathogenicity rating`
```

```{r}
histograms$`COGEM pathogenicity rating`
```

```{r}
boxplots$`COGEM pathogenicity rating`
```

# Other datasets

+ Some datasets contain values different to logical or numeric in the
"Attribute_value" column.

```{r}
other_datasets <- phys %>% 
    discard(~ is.logical(.x$Attribute_value) || is.numeric(.x$Attribute_value))

for (i in seq_along(other_datasets)) {
    other_datasets[[i]]$NCBI_ID <- as.character(other_datasets[[i]]$NCBI_ID)
}

other_datasets <- map(other_datasets, ~ left_join(.x, taxonomy_table, by = "NCBI_ID"))
other_datasets_names <- names(other_datasets)
other_datasets_names
```

```{r}
other_datasets_summary_tables <- map(other_datasets_names, ~ attributeSummaryByRank(other_datasets, .x, categorical = FALSE)) %>% 
    set_names(other_datasets_names)
```

## Acetate producing

```{r}
other_datasets_summary_tables$`acetate producing`
```

## Lactate producing

```{r}
other_datasets_summary_tables$`lactate producing`
```

## Width

```{r}
other_datasets_summary_tables$width
```

## Hydrogen gas producing

```{r}
other_datasets_summary_tables$`hydrogen gas producing`
```

## Length

```{r}
other_datasets_summary_tables$length
```

# Fatty acid composition

+ Dataset imported with the `fattyAcidComposition` function.

```{r}
fac <- phys["fatty acid composition"]
fac[[1]] <- left_join(fac[[1]], taxonomy_table, by = "NCBI_ID")
names(fac) <- "fatty acid composition"
```

```{r}
fac_summary_table <- attributeSummaryByRank(fac, "fatty acid composition", categorical = FALSE)
```

```{r}
fac_summary_table
```

```{r}
histograms$`fatty acid composition`
```

```{r}
boxplots$`fatty acid composition`
```

```{r}
sessioninfo::session_info()
```

